\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{enumitem}

\newcommand{\defto}{\Rightarrow}
\newcommand{\strictto}{\rightarrow}
\newcommand{\defeatto}{\leadsto}

\pagestyle{empty}

\begin{document}

\begin{center}
{\Large\bfseries Defeasible Abduction for Mathematical Discovery}
\end{center}

\bigskip

\noindent
Lakatos showed that mathematical discovery is defeasible: conjectures
are revisable defaults, counterexamples are defeaters, and conjecture
refinement is conservative belief revision. We propose applying the
Defeasible Abduction Benchmark (DeFAb) framework to mathematics,
producing a benchmark (\textsc{DeFAb-Math}) that tests the full
conjecture--counterexample--refinement cycle using Lean~4 as an exact
verifier. Instances are generated from Mathlib (193K+ formalizations),
MMLKG (65K+ Mizar theorems), and Sequencelib/OEIS (25K+ formalized
sequences). All outputs are open-source.

\bigskip

\section*{The Isomorphism}

A DeFAb defeasible theory has facts, strict rules, defeasible rules,
defeaters, and a superiority relation. Formalized mathematics provides
the same structure:

\begin{itemize}[nosep]
\item \textbf{Strict rules} from proven theorems.
  \texttt{theorem foo (h1:~C1) (h2:~C2) : P} becomes
  $r_s\colon C_1, C_2 \strictto P$, validated by Lean's kernel.

\item \textbf{Defeasible rules} from weakened theorems. Dropping
  condition $C_2$ yields $r_d\colon C_1 \defto P$, a statement that is
  usually true but admits exceptions when $C_1 \wedge \neg C_2$ holds.

\item \textbf{Defeaters} from counterexamples. An instance $a$ with
  $C_1(a) \wedge \neg C_2(a) \wedge \neg P(a)$ is a defeater
  $r^*$ with $r^* \succ r_d$. Lean verifies both sides.

\item \textbf{Conservativity} from the full theorem. The defeater must
  block $P$ only in the exception region $(\neg C_2)$, preserving
  every instance where all original conditions hold.
\end{itemize}

\section*{Three Levels}

DeFAb generates instances by ablating critical elements from a complete
theory. In \textsc{DeFAb-Math}:

\smallskip
\noindent\textbf{Level 1} (missing lemma): remove a fact; the model
identifies which known result completes the proof.

\noindent\textbf{Level 2} (theorem abduction): remove a rule; the model
reconstructs the theorem statement, including the correct conditions.

\noindent\textbf{Level 3} (counterexample construction): remove the
defeater; the theory now makes an incorrect default prediction, and the
model must construct a minimal counterexample that blocks it
\emph{conservatively}, preserving all correct instances.

\smallskip
\noindent\textit{Example.} Euler's formula $V{-}E{+}F=2$. The
defeasible rule applies to all polyhedra. The defeater: genus $> 0$
(torus has $V{-}E{+}F=0$). At Level~3 the model must construct the
genus-based exception without breaking the formula for convex polyhedra.

\noindent\textit{Example.} ``Convergent sequences of continuous
functions have continuous limits.'' The defeater: pointwise (not
uniform) convergence may yield discontinuous limits. The refined
conjecture is a strict rule in Mathlib
(\texttt{TendstoUniformly.continuous}).

\noindent\textit{Example.} ``Groups of prime-power order are abelian.''
The defeater: order $p^3$ admits non-abelian groups (Heisenberg group
over $\mathbb{F}_p$). Conservative refinement: order $p^2$ groups are
abelian.

\section*{Extraction Pipeline}

\begin{enumerate}[nosep]
\item \textbf{Ingest} Mathlib as a typed dependency graph.
\item \textbf{Synthesize defeasible rules} by dropping hypothesis
  subsets; filter to retain variants that are both non-trivially true
  and non-trivially defeasible.
\item \textbf{Harvest counterexamples} via Mathlib mining, Lean's
  \texttt{decide} tactic, and LLM-assisted generation verified by Lean.
\item \textbf{Generate instances} at all three levels using the DeFAb
  ablation procedure, with Lean as the verifier at every stage.
\end{enumerate}

\section*{Verifier Advantage}

Lean's type checker replaces DeFAb's polynomial-time defeasible
derivation checker with an exact, sound, and complete oracle. Gold
standards are verified without approximation. The DeFAb
verifier-in-the-loop RLHF pipeline transfers directly: Lean serves as
the reward function for DPO and PPO, eliminating reward model error.
The five-tier graded scoring (well-formedness $\to$ anomaly resolution
$\to$ conservativity $\to$ strength) applies unchanged.

\section*{Tautological Bottleneck}

The most valuable conjectures are absent from training data because,
once proven, they become established facts. \textsc{DeFAb-Math} works
backwards: weaken a proven theorem to produce a false conjecture the
model has never seen, then ask it to discover the counterexample. This
generates unbounded novel instances from existing formalizations, each
with a Lean-verified gold standard. Unlike existing counterexample
benchmarks, the task includes constructing a \emph{conservative
refinement} that preserves all correct instances.

\section*{What Transfers from DeFAb}

All formal definitions (defeasible theories, support sets, criticality,
conservativity, revision distance, graded scoring), the four-modality
evaluation harness, the three-stage decoder, the error taxonomy
(E1--E5), and the DPO/VITL-RLHF fine-tuning pipeline. The new
engineering is the Mathlib extraction pipeline and the Lean verifier
integration.

\end{document}
