\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{titlesec}
\usepackage{tikz}
\usetikzlibrary{positioning, arrows.meta, calc}

\titlespacing*{\section}{0pt}{8pt}{4pt}
\titlespacing*{\subsection}{0pt}{6pt}{3pt}
\setlength{\parskip}{4pt}
\setlength{\parindent}{0pt}
\pagestyle{empty}

\begin{document}

\begin{center}
{\Large\bfseries Verifier-Backed Defeasible Fine-Tuning}\\[2pt]
{\normalsize Targeting the Three Structural Deficits of Foundation Models}
\end{center}

\vspace{-3pt}\hrule\vspace{4pt}

\section*{The Three Deficits}

Current foundation models suffer three entangled structural deficits. Each is well-documented; together they account for the failures that limit deployment.

\begin{enumerate}[nosep, leftmargin=1.6em, label=\textbf{\arabic*.}]
    \item \textbf{Grounding.} Models cannot trace conclusions to supporting evidence. They confuse strict knowledge with revisable defaults. This is the root cause of hallucination.
    \item \textbf{Novelty.} Without knowing which beliefs are revisable, models cannot identify where creative exceptions might apply. This is the root cause of creative sterility on out-of-distribution tasks.
    \item \textbf{Belief revision.} Models cannot update knowledge while preserving unrelated commitments. Editing one fact triggers cascading side effects (the ripple problem documented for ROME, MEMIT, and successors).
\end{enumerate}

The three deficits share a structural cause: the absence of an explicit \emph{defeasible knowledge structure} that marks rules as strict or revisable, traces every conclusion to its support set, and admits a polynomial-time verifier that decides whether a proposed update preserves unrelated conclusions (the formal counterpart of AGM minimal change). The project supplies this structure and uses it to fine-tune foundation models against the three deficits directly.

\section*{The Fine-Tuning Pipeline}

\begin{center}
\begin{tikzpicture}[
    >=Latex,
    box/.style={
        draw, rounded corners=3pt,
        minimum height=0.85cm, minimum width=2.4cm,
        font=\footnotesize, inner sep=4pt,
        text centered, align=center
    },
    fixed/.style={
        draw, very thick, rounded corners=3pt,
        minimum height=0.85cm, minimum width=2.6cm,
        font=\footnotesize\bfseries, inner sep=4pt,
        text centered, align=center
    },
    arr/.style={->, thick},
    lbl/.style={font=\scriptsize, inner sep=2pt}
]
    \node[box] (sft) at (0, 0) {Phase 1: SFT\\gold hypotheses};
    \node[box] (dpo) at (3.4, 0) {Phase 2: DPO\\margin-weighted};
    \node[box] (grpo) at (6.8, 0) {Phase 3: GRPO\\self-play};
    \node[box] (az) at (10.2, 0) {Phase 4: AlphaZero\\expert iteration};

    \node[fixed] (ver) at (5.1, -2.3) {Polynomial-time verifier\\(fixed point, never drifts)};
    \node[box, dashed] (kb) at (10.4, -2.3) {Knowledge base\\$+$ live environment};

    \draw[arr] (sft) -- node[lbl, above] {warm} (dpo);
    \draw[arr] (dpo) -- node[lbl, above] {pref} (grpo);
    \draw[arr] (grpo) -- node[lbl, above] {policy} (az);

    \draw[arr] (ver.north -| sft.south) -- (sft.south) node[lbl, midway, right] {targets};
    \draw[arr] (ver.north -| dpo.south) -- (dpo.south) node[lbl, midway, right] {scored pairs};
    \draw[arr] (ver.north -| grpo.south) -- (grpo.south) node[lbl, midway, right] {rewards};
    \draw[arr] (ver.north -| az.south) -- (az.south) node[lbl, midway, right] {tree returns};

    \draw[arr] (kb.west) -- node[lbl, above] {games} (ver.east);
    \draw[arr, dashed] (az.east) -- ++(0.6, 0) |- node[lbl, right, pos=0.25] {harder games} (kb.east);
\end{tikzpicture}
\end{center}

\vspace{-2pt}
{\footnotesize Figure~1. The same verifier scores every stage. Each phase consumes a different shape of verifier output: gold targets, scored preference pairs, group-relative rewards, and tree-search returns. No learned reward model anywhere in the loop.}

\section*{How Each Deficit Is Targeted (and Why We Expect Success)}

\subsection*{1. Grounding}

\textbf{Mechanism.} Every conclusion in a defeasible theory has a traceable support set. Phase~1 (SFT) trains models on verified gold hypotheses at fact completion (L1) and rule abduction (L2): the model learns to operate over the support structure rather than retrieve from the training distribution. Phase~2 (margin-weighted DPO) reinforces the format and the structural mapping with preference pairs scored exactly by the verifier.

\textbf{Why we expect success.} The capability is already near-ceiling under direct prompting: the four-model frontier panel achieves 73--79\% on L2 across formal modalities at baseline. The gap to close is narrow, the targets are formally verified rather than crowd-labelled, and the structural form of the task matches the training signal exactly. Fine-tuning gains here are the most predictable of the three.

\subsection*{2. Novelty}

\textbf{Mechanism.} Phase~3 (GRPO self-play) and Phase~4 (AlphaZero expert iteration) operate on defeater abduction (L3): the model must construct exception rules in a controlled novel vocabulary. Self-play generates an automatic curriculum (improvements on one side of a defeasible conflict force improvements on the other), and neural-guided MCTS extends the search horizon beyond what one-shot generation can reach. The synthetic-contamination control guarantees the answer is not in the training corpus, so the model is forced to construct rather than retrieve.

\textbf{Why we expect success.} AlphaProof (2025) demonstrated that neural-guided MCTS with an exact verifier reaches silver-medal IMO performance starting from a domain (formal mathematics) where the verifier is the proof kernel. DeepSearch (ICLR 2026) demonstrated that tree-search-augmented RLVR overcomes the well-documented GRPO plateau. Defeasible reasoning admits the same exact, polynomial-time verifier that the AlphaGo lineage requires, and is one of the very few non-mathematical domains that does so. The recipe transfers because the verifier transfers.

\subsection*{3. Belief Revision}

\textbf{Mechanism.} The conservativity check (formal counterpart of AGM minimal change) is computed in polynomial time by the verifier. Margin-weighted DPO weights each preference pair by the verifier-scored conservativity gap, so the loss puts gradient on the qualitative transitions (non-resolutive output to conservative resolution) rather than on cosmetic differences. GRPO self-play uses the same conservativity score as exact reward; AlphaZero-style expert iteration scores tree leaves by it directly. ROME, MEMIT, and other knowledge-editing methods cannot do this because they lack any measure of unrelated-conclusion preservation; the project supplies that measure as a training signal.

\textbf{Why we expect success.} The capability is latent rather than absent: chain-of-thought scaffolding lifts L3 accuracy by 56--79 percentage points for reasoning-optimised models, confirming the model can revise correctly when the structure is supplied externally. Verifier-backed fine-tuning supplies the structure internally. The pre-registered hard subset (stratified along novelty, depth, multi-anomaly conservativity, and synthetic contamination) provides the non-saturated evaluation surface needed to measure genuine fine-tuning lift rather than memorisation.

\section*{Practical Applications}

The recipe is domain-general; only the knowledge base changes.

\begin{itemize}[nosep, leftmargin=1.4em]
    \item \textbf{Biomedical decision support.} Drug, comorbidity, and pharmacogenomic exception reasoning over UMLS and SNOMED~CT, where confident default-following is the clinically dangerous failure mode.
    \item \textbf{Functional genomics.} Curated negative annotations become first-class fine-tuning signal instead of being silently dropped.
    \item \textbf{Regulatory and legal reasoning.} Conservativity guarantees a fine-tuned compliance agent does not silently invalidate unrelated rulings when one rule changes.
    \item \textbf{Materials and chemistry discovery.} Conjecture refinement with verifier-checked guardrails against false-default predictions.
    \item \textbf{Formal mathematics and program verification.} Same pipeline with Lean's kernel as verifier and Mathlib as the knowledge base.
    \item \textbf{Safety-critical autonomous agents.} Robotics, unmanned systems, and operator-in-the-loop control, demonstrated end-to-end on a live StarCraft~II bridge under three enforcement modes (trust, audit, verifier-gated re-prompt).
\end{itemize}

\section*{Why Now}

Verifier-backed RL is the most tractable known route past reward hacking. Defeasible logic is one of the few non-mathematical domains that supplies the polynomial-time exact verifier the recipe requires. The full stack is built, MIT-licensed, and reproducible on commodity HPC.

\end{document}
